Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-14 paper/paper.tex
sum proof wip

\section{Denotational Semantics}
\steven{Define a Kleene, Lambek, and Chomsky category}

Let $\semcat$ be a Chomsky category.

\steven{Finish the denotation of lin types figure for spf, el , map , mu}
\begin{figure}
  \begin{align*}
    \semg{I} &= I \\
    \semg{A \otimes B} &= \semg{A} \otimes \semg{B} \\
    \semg{A \lto B}  &= \semg{A} \lto \semg{B} \\
    \semg{B \tol A} &= \semg{B} \tol \semg{A} \\
    \semg{\bigoplus_{x:X} A} &= \coprod_{x:X} \semg{A} \\
    \semg{\bigwith_{x:X} A} &= \prod_{x : X} \semg{A} \\
    \semg{\uparrow A} &= Hom(I,A) \\
    \semg{\equalizer{a:A}{f}{g}} &= Eq\left( \semg{f} , \semg{g} \right) \\
    \sem{L}\gamma &= \semcat_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)
  \end{align*}
  \caption{Denotation of Linear Types}
\end{figure}

@@

\subsubsection{$\bigamp$-Products}
% Products
\paragraph{$\bigamp$-Introduction}
$\semg{\dlamb x e} : \semg{\Delta} \to \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}$ is
defined by the universal property of the product
\[
\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
\]

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
  \sem{A}(\gamma , M)
\end{tikzcd}
\end{center}

@@

\subsubsection{$\bigoplus$-Sums}
% Sums
\paragraph{$\bigoplus$-Introduction}
$\semg{\sigma~M~e} : \semg{\Delta} \to \LinSigTyLimit {x} {X} \sem{A}(\gamma, x)$

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
  \LinPiTyLimit {x} {X} {\sem{A}(\gamma , x)}
\end{tikzcd}
\end{center}

@@

\begin{center}
% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
\begin{tikzcd}[column sep =large]
{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
{\left( \semg{\Delta'_1} \otimes \LinSigTyLimit {x} {X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
{\LinSigTyLimit {x} {X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\LinSigTyLimit {x} {X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
\end{tikzcd}
\end{center}

@@

\paragraph{$\bigoplus\beta$}
The $\beta$ rule for $\bigoplus$ is given by,
\[
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
\]
\begin{proof}
  \begin{align*}
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes i_M \circ \semg{a'}) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes i_M \circ \lambda) \otimes id \circ m \otimes id \circ m
  \end{align*}
\steven{TODO need to track subscripts on $m$ in the $\bigoplus\beta$ proof, and
  finish this part of the above part of the proof}

  \begin{align*}
    \semg{e'[M/x,a'/a]}
    &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
    &= \semg{e'}
  \end{align*}
  \steven{TODO shouldn't the semantics of $e'$ be in context $\gamma,x$ not $x$?}
\end{proof}

>>>>

\section{Denotational Semantics}
Here we extend the denotational semantics from \cref{sec:semantics-and-metatheory} to cover all of \theoryname syntax.

@@

\subsubsection{$\bigamp$-Products}
% Products
\paragraph{$\bigamp$-Introduction}
$\semg{\dlamb x e} : \semg{\Delta} \to \prod_{x : X} {\sem{A}(\gamma , x)}$ is
defined by the universal property of the product
\[
\semg{\dlamb x e} = \left( \sem{e}(\gamma , x) \right)_{(x : X)}
\]

@@

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \prod_{x : X} {\sem{A}(\gamma , x)} \arrow[r , "\pi_M"] &
  \sem{A}(\gamma , M)
\end{tikzcd}
\end{center}

@@

\subsubsection{$\bigoplus$-Sums}
% Sums
\paragraph{$\bigoplus$-Introduction}
$\semg{\sigma~M~e} : \semg{\Delta} \to \coprod_{x : X} \sem{A}(\gamma, x)$

\begin{center}
\begin{tikzcd}
  \semg{\Delta} \arrow[r , "\semg{e}"] &
  \sem{A}(\gamma, M) \arrow[r , "i_M"] &
  \coprod_{x : X} {\sem{A}(\gamma , x)}
\end{tikzcd}
\end{center}

@@

\begin{center}
% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAB124YBbAc2CcAIjAY46AcgD6ARlLDR4+exFjJUgEwBfEFtLpMufIRRzK1es1aIOXXgIVrpcx+K0ACThDw94nu-yCKorq2rr6IBjYeAREcjJUtIwsbJwMMABmOAAU-tyBruoyHl4+fpz5DsFqJewATlh8ABY4AJT+3li+cHn2Qari0mF6BtHGcaQaiZYpNpwAMlhgAMqNACoAnos8uO7AAB4ewAAaRxW8-SHOpHTuiO4AgsoDoVrZnHx0PDy3pO77rR0I0ihhiJhIk2myWstkWK3WWy6uwOR1OezSmRyvR4lycsjeHy+P3cfwBHTKcHYYHcNNp5xxDwJ7E+31+-3anAazTa5K65QCuMGmiZLOJpMB4VGRliKDICQs0NS7HSWVy9KqL2ctU63X8cNWfE222Rhz2aKC+WAjPezKJbIBtS5LQ57B1-MqgtekpBYxlyDkAGYoVYlZaAMIiu3i71RaXgshBhUhuYBAQR3SJGBQPjwIigDJ1CA8JBkEA4CBIORJZMgHhSYA2zWyZ4hVotvFhagMOgAI1EAAVQeMbCqcN6C0XK9Ry0gNEnZrX64VroUtKU+T0sFAQF3ewOhzKQKPx4Xi4g52WK4gAKzzmHZLe83Xq4AwLQut2b7e7vsMQe+kwj0xE9J0QANpyvAA2O82GQbFXwkSNWRJdl3Aoetsn2O53GOCUf33AC2GPYEJzPAAWCCkHA6sFx4AA9YAAFpigw5dmzoRAHjbNjOyPPc-wPQDiIiUikFvS8kAomiYR3Pjf3-OMiOAkjTyQaCJMQAB2LQKC0IA
\begin{tikzcd}[column sep =large]
{\semg{\Delta'_1,\Delta,\Delta'_2}} \arrow[r, "{m_{(\Delta'_1,\Delta),\Delta'_2}}"]                                                                                                           & {\semg{\Delta'_1,\Delta} \otimes \semg{\Delta'_2}} \arrow[d, "{m_{\Delta'_1,\Delta}\otimes id}"]                             \\
{\left( \semg{\Delta'_1} \otimes \coprod_{x:X} {\sem{A}(\gamma , x)} \right) \otimes \semg{\Delta'_2}} \arrow[d, "d"]                                                                 & \left( \semg{\Delta'_1} \otimes \semg{\Delta} \right) \otimes \semg{\Delta'_2} \arrow[l, "(id \otimes \semg{e}) \otimes id"] \\
{\coprod_{x : X} {\left( \sem{\Delta'_1}(\gamma , x) \otimes
      \sem{A}(\gamma , x) \right) \otimes \sem{\Delta'_2}(\gamma , x)}} \arrow[r, "{m^{-1}_{(\Delta'_1,a:A),\Delta'_2}}"] & {\coprod_{x : X} {\sem{\Delta'_1,a : A,\Delta'_2}(\gamma , x)}} \arrow[d, "{[ \sem{e'}(\gamma , x) ]_{(x : X)}}"]     \\
\semg{C}                                                                                                                                                                                      & {\sem{C}(\gamma, x)} \arrow[l]
\end{tikzcd}
\end{center}

@@

\paragraph{$\bigoplus\beta$}
The $\beta$ rule for $\bigoplus$ is given by,
\[
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}} = \semg{e'[M/x,a'/a]}
\]
\begin{proof}
  \begin{align*}
  \semg{\letin {\sigma~x~a} {\sigma~M~a'} {e'}}
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes \semg{\sigma~M~a'}) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes (i_M \circ \semg{a'})) \otimes id \circ m \otimes id \circ m \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ m^{-1} \circ d \circ (id
  \otimes (i_M \circ \lambda)) \otimes id \circ m \otimes id \circ m \\
  & = ? \\
  & = {[ \sem{e'}(\gamma , x) ]_{(x : X)}} \circ i_{M} \\
  & = \sem{e'}(\gamma , x)
  \end{align*}

  \steven{How does the distributivity morphism handled? Seems like its a barrier to invoking coherence}

  \begin{align*}
    \semg{e'[M/x,a'/a]}
    &= \semg{e'} \circ \semg{M/x,a'/a} \tag{\cref{lem:subst}} \\
    &= \sem{e'}(\gamma , x)
  \end{align*}
\end{proof}
